---
title: "The Digit Recognizer Project"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
```

## The project

Through the Optimistic Data Spirit, this is my process in R for the Kaggle competition to learn computer vision fundamentals. For all details about this competition see <https://www.kaggle.com/c/digit-recognizer/data>.

## The data

The MNIST database (Modified National Institute of Standards and Technology database) is a large database of handwritten digits that is commonly used for training various image processing systems (Wikipedia). Each image is 28 pixels in height and 28 pixels in width, for a total of 784 pixels. Each pixel has a single pixel-value between 0 and 255 associated with it, indicating the lightness or darkness of that pixel, with higher numbers meaning darker (Kaggle website).

We received three csv files: 

* train.csv - the training dataset with 785 columns. The first column, called "label", is the digit that was drawn by the user. The rest of the columns contain the pixel-values of the associated image. Each pixel column in the training set has a name like pixelx, where x is an integer between 0 and 783, inclusive. To locate this pixel on the image, suppose that we have decomposed x as x = i * 28 + j, where i and j are integers between 0 and 27, inclusive. Then pixelx is located on row i and column j of a 28 x 28 matrix, (indexing by zero).

* test.csv - same as the training set, except that it does not contain the "label" column.

* sample-submission.csv - for each of the 28000 images in the test set, output a single line containing the ImageId and the digit you predict.

## The evaluation

The evaluation metric for this contest is the categorization accuracy, or the proportion of test images that are correctly classified. For example, a categorization accuracy of 0.97 indicates that you have correctly classified all but 3% of the images.



```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
